Cluster Dynamics with HETDEX at z < 0.5 - I: Simulated Performance, Mass Recovery, and Limits to Cosmology

### 2016-09-26 paper1.tex
update paper
new figures and some tweaked text

The second broad method is the direct calibration of cluster masses. This recipe has two distinct but not always independent tracks. The ``direct'' method uses observations of a relatively small set of clusters and then uses observable mass proxies, including X-ray temperatures and luminosities \citeeg{Mantz2010, Rykoff2014}, microwave observations \citeeg{Vanderlinde2010, Sehgal2011}, optical richness \citeeg{Abell1958, Rykoff2012} or weak lensing (WL; \eg\ \citealt{Rozo2010}) as examples, which provide a ``true'' mass. This directly calibrates the observable-mass relation which is then applied to a much larger sample. The complications lie in that the ``true'' masses are, in fact, estimations, and the methods used to recover these cluster masses are subject to their own limitations. X-ray based cluster masses assume hydrostatic equilibrium \citeeg{Mantz2015} which may only be valid for a very small number and range of cluster masses. The Sunyaev-Zel'dovich Effect (SZE; \citealt{Sunyaev1972}), which uses the up-scattering of cosmic microwave background (CMB) photons to estimate cluster masses, provides accurate estimations of mass, but the ability to detect low mass galaxy clusters is currently limited by technology \citeeg{Carlstrom2002a} and can also be effected by the properties of the intracluster medium \citeeg{Pipino2010}. WL estimates are, in principle, correct in the mean, but they suffer from signal-to-noise requirements, limiting their usefulness in low mass clusters (where the lensing signal is particularly weak), and potentially suffer from line-of-sight effects as WL is sensitive to all mass along the line-of-sight. Virial mass estimators which determine the cluster mass based on the motions of the member galaxies \citeeg{Ruel2014, Sifon2015} are promising in that it is a direct measurement of the depth of clusters potential well, but suffer from systematics due to cluster formation physics which disrupts the velocity field.

@@

\subsection{HETDEX}\label{sec: hetdex}
\begin{figure}
\includegraphics[width=\columnwidth]{figures/pointings.pdf} 
	\caption[HETDEX pointing strategy.]{Representative observation tiling scheme for the HETDEX $16' \times 16'$ pointings. Each turquoise square represents the position of a single VIRUS IFU, and the dashed octagons approximate the size of a single observation. Inside each IFU HETDEX will achieve near complete coverage through three dithers. See the text for more details.} 
	\label{fig: ifu layout} 
\end{figure}

@@

\subsection{Mock Observations}\label{sec: observations}
When selecting galaxies from the Buzzard catalog we assume an observation for all galaxies laying within a turquoise, IFU square in Figure~\ref{fig: ifu layout}. In practice, this is achieved by three dither positions at each pointing. Galaxies which lie between the IFUs are missed, as well as the galaxies which lie between the pointings, as there is no overlap between one pointing and the next. To cover the 398.49 \degsq\ field of the Buzzard catalog we require 5370 pointings where 0.015 \degsq\ of each pointing is covered by an IFU. The total area of the sky covered by an IFU is 80.80 \degsq\ which gives a filling factor of 1/4.65, slightly decreased from the expected filling factor of 1/4.5.

@@

\section{Recovery of Parameters}\label{sec:recovery}
\subsection{Cluster Redshift}
The accurate determination of the cluster redshift ($z_c$) is crucial to the reliability of all following measurements. An incorrect cluster redshift introduces error into the measured line-of-sight velocity (LOSV) and corresponding dispersion, which, in turn, contributes to errors associated with dynamical mass and radius.

In simple terms, the cluster redshift is the mean of the redshifts of all galaxies associated with the cluster, where the mean is the first moment of the velocity (redshift) distribution function $P(z)$. In practice, the first moment is strongly subject to outliers, so we rely instead on the biweight location estimator \citep{Beers1990} through\footnote{Implemented as part of the {\sc astLib} Python library. See \hbox{\url{http://astlib.sourceforge.net}}}:
\begin{equation}
z_c = z_{med} + \frac{\sum_{|u_i| < 1} (z_i - M)(1-u_i^2)^2}{\sum_{|u_i| < 1}(1-u_i^2)^2}
\end{equation}
where $z_i$ are the individual redshifts, $z_{med}$ is the median redshift and $u_i$ is given by:
\begin{equation}
	u_i = \frac{(z_i - M)}{C\,\mathrm{MAD}}.
\end{equation}
MAD is the median absolute deviation, also defined in \cite{Beers1990}, and $C$ is the a tuning constant. We choose $C=6$ (the suggested value) which balances computational speed and location accuracy. 
% Although this work assumes that we know each galaxy's redshift to infinite precision, in practice, we find a simple weighted mean provides a reliable estimate of $z_c$ when there are uncertainties on the individual galaxy redshifts.

@@

For the clusters making up the \emph{test} sample the mass is unknown (it is what we are trying to predict) but the other features are known. To determine the mass probability distribution of a test cluster, $P(\log\ M_{200c})$, we combine the conditional probability distribution, $P(\log\ M_{200c}|\vec{x})$, created previously with the probability distribution of $\sigma$, the LOSVD, through Equation~\ref{eq: Pm}.
\begin{equation}\label{eq: Pm}
	P(\log\ M_{200c}) = \int P(\log\ M_{200c}|\vec{x})\ P(\log\ \sigma)\ d\log\ \sigma\ P(z)\ dz
\end{equation}
The expected mass is determined by calculating the first moment of the probability density. This becomes our ``predicted'' cluster mass, $M_{pred}$.
\begin{equation}\label{eq: expected mass}
	\log\ M_{pred}= \int \log\ M_{200c}\ P(\log\ M_{200c})\ d\log\ M_{200c}
\end{equation}
The confidence interval associated with this prediction can be estimated two ways. First, by calculating the second moment of the probability density through
\begin{equation}\label{eq: variance}
	\mathrm{Var} = \int (\log\ M_{200c} - \log\ M_{pred})^2\ P(\log\ M_{200c})\ d\log\ M_{200c}
\end{equation}
or by drawing many samples from $P(\log\ M_{200c})$ and calculating the values at the 16th and 84th percentile. In practice we find that both methods produce similar results for a large number of trials. Therefore, we quote predicted masses as the most probable mass given by Equation~\ref{eq: expected mass} and associated 68\% error estimated through the square root of Equation~\ref{eq: variance}.

@@

Because of the simplifying assumptions, and the superior quality of the data (no contamination, signal-to-noise issues, etc.), realistic expectations for HETDEX is to directly constrain $\sigma_8$ is not yet competitive with other methods (\eg, CMB, WL, X-ray). For example, \cite{DeHaan2016} constrain $\sigma_8$ to $\sim5$\% using a sample of 337 SZE detected clusters from the SPT-SZE survey. For the $\sim1500$ clusters detected with Survey observations to constrain $\sigma_8$ and to be dominated by cluster statistics alone ($\Delta \ln N \sim N^{-1/2}$), the absolute cluster calibration would need to be better than 2.5\%. For a fully Targeted survey, about 14,000 clusters, this cluster mass calibration uncertainty reduces to $>1\%$. Nevertheless, while analysis of HETDEX clusters may not provide the most accurate constraint on $\sigma_8$, it will provide important systematics checks on other studies and will ultimately improve the measurements of $\sigma_8$.

@@

We find that we are able to accurately recover an average intrinsic scatter of $0.2 <\langle \sigma_{M|\lambda} \rangle <0.3$ dex, finding $\langle \sigma_{M|\lambda} \rangle = 0.257\pm0.007$ at $\sigma_{true} = 0.25$ with Survey observations. This is very promising as other observational studies have estimated the intrinsic scatter of real clusters to be $\sim0.25$ dex \citeeg{Rozo2014, Rozo2015}. As the intrinsic scatter increases or decreases, we fail to recover the scatter as accurately.

For the richness range $10 \leq \lambda < 130$ the intrinsic scatter in between the $\mathrm{M_{pred,corr}}$ predicted cluster masses and the true cluster mass (the basis of our richnesses) is $\sim0.15$ dex for Targeted observations and $\sim0.20$ dex for Survey observations. This forms an effective floor on the amount of scatter we are able to recover from our fitting process. As we reduce the overall scatter in our cluster mass recovery, this floor will lower. We underestimate the scatter in high intrinsic scatter relations because any residual bias remaining in the predicted clusters masses after correction reduces the observed scatter. Because the bias subtraction when creating $\mathrm{M_{pred,corr}}$ subtracts the mean bias, we are left with a small amount of residual bias, lowering the measured scatter for high scatter relationships.

\section{Summary}\label{sec:summary}
Here, we present detailed simulations of the upcoming HETDEX survey's applicability to the detection and total mass measurement of galaxy clusters based on emission line and continuum redshifts for galaxies at z < 0.5. Using mock galaxy catalogs and HETDEX-like observational strategies and limits, we observe our simulated sky, estimate the number of clusters observed, and derive basic cluster parameters, redshift, line-of-sight velocity dispersion (LOSVD). Using a traditional power law-based, velocity dispersion, scaling relation along with more advanced probability and machine learning (ML) techniques, we estimate the total mass for each cluster. We discuss each cluster mass estimate's precision, and discuss the ability of HETDEX to constrain the cosmological parameter $\sigma_8$ based on those predicted cluster masses. In addition, we comment on how HETDEX may improve current and future photometric large-area sky surveys' cluster mass estimates derived from optical richness.

>>>>

The second broad method is the direct calibration of cluster masses. This recipe has two distinct but not always independent tracks. The ``direct'' method uses observations of a relatively small set of clusters and then uses observable mass proxies, including X-ray temperatures and luminosities \citeeg{Mantz2010, Rykoff2014}, microwave observations \citeeg{Vanderlinde2010, Sehgal2011}, optical richness \citeeg{Abell1958, Rykoff2012} or weak lensing (WL; \eg\ \citealt{Rozo2010}) as examples, which provide a ``true'' mass. This directly calibrates the observable-mass relation which is then applied to a much larger sample. The complications lie in that the ``true'' masses are, in fact, estimations, and the methods used to recover these cluster masses are subject to their own limitations. X-ray based cluster masses assume hydrostatic equilibrium \citeeg{Mantz2015} which may only be valid for a very small number and range of cluster masses. The Sunyaev-Zel'dovich Effect (SZE; \citealt{Sunyaev1972}), which uses the up-scattering of cosmic microwave background (CMB) photons to estimate cluster masses, provides accurate estimations of mass, but the ability to detect low mass galaxy clusters is currently limited by technology \citeeg{Carlstrom2002a} and can also be effected by the properties of the intracluster medium \citeeg{Pipino2010}. WL estimates are, in principle, correct in the mean, but they suffer from signal-to-noise requirements, limiting their usefulness in low mass clusters (where the lensing signal is particularly weak), and potentially suffer from line-of-sight effects as WL is sensitive to all mass along the line-of-sight. Virial mass estimators which determine the cluster mass based on the motions of the member galaxies \citeeg{Ruel2014, Sifon2015} are promising in that it is a direct measurement of the depth of clusters potential well, but suffer from systematics due to cluster formation physics which disrupts the velocity field, such as cluster mergers \citeeg{Evrard2008} and tidal stripping \citeeg{Faltenbacher2006}.

@@

\subsection{HETDEX}\label{sec: hetdex}
\begin{figure}
\includegraphics[width=\columnwidth]{figures/pointings.pdf} 
	\caption[HETDEX pointing strategy.]{\emph{Top:} Representative observation tiling scheme for the HETDEX $16' \times 16'$ pointings over a small region of the sky. \emph{Bottom:} The simplified tiling pattern used in this study. In both schemes, the dashed octagons approximate the size of a single observation. Each observation consists of the 78 VIRUS IFUs, represented by the turquoise squares in the bottom panel. Inside each IFU HETDEX will achieve near complete coverage through three dithers. See the text for more details.} 
	\label{fig: ifu layout} 
\end{figure}

@@

\subsection{Mock Observations}\label{sec: observations}
When selecting galaxies from the Buzzard catalog we assume an observation for all galaxies laying within a turquoise, IFU square in Figure~\ref{fig: ifu layout}. In practice, this is achieved by three dither positions at each pointing. Galaxies which lie between the IFUs are missed, as well as the galaxies which lie between the pointings, as there is no overlap between one pointing and the next. We use a pointing pattern which differs slightly from the expected final HETDEX observing pattern. See Figure~\ref{fig: ifu layout} for an illustration of the difference. To cover the 398.49 \degsq\ field of the Buzzard catalog we require 5370 pointings where 0.015 \degsq\ of each pointing is covered by an IFU. The total area of the sky covered by an IFU is 80.80 \degsq\ which gives a filling factor of 1/4.65, slightly (about 3\%) decreased from the expected filling factor of 1/4.5.

@@

\section{Recovery of Parameters}\label{sec:recovery}
In the following sections, we assume that we know a priori which galaxies belong to each individual cluster. So while our observations may include many field galaxies which are not belonging to the cluster in question, we exclude those galaxies during the analysis of the cluster parameters. Cluster membership is a critical issue which must be addressed when utilizing real observations. To that end, we assay a small sample of real galaxy clusters in a follow up work (Boada \etal\ in Preparation).

\subsection{Cluster Redshift}
The accurate determination of the cluster redshift ($z_c$) is crucial to the reliability of all following measurements. An incorrect cluster redshift introduces error into the measured line-of-sight velocity (LOSV) and corresponding dispersion, which, in turn, contributes to errors associated with dynamical mass and radius.

In simple terms, the cluster redshift is the mean of the redshifts of all galaxies associated with the cluster, where the mean is the first moment of the velocity (redshift) distribution function $P(z)$. In practice, the first moment is strongly subject to outliers, so we rely instead on the biweight location estimator \citep{Beers1990} through\footnote{Implemented as part of the {\sc astLib} Python library \citep{astlib}. See \hbox{\url{http://astlib.sourceforge.net}}}:
\begin{equation}
z_c = z_{med} + \frac{\sum_{|u_i| < 1} (z_i - M)(1-u_i^2)^2}{\sum_{|u_i| < 1}(1-u_i^2)^2}
\end{equation}
where $z_i$ are the individual redshifts, $z_{med}$ is the median redshift and $u_i$ is given by:
\begin{equation}
	u_i = \frac{(z_i - M)}{C\,\mathrm{MAD}}.
\end{equation}
MAD is the median absolute deviation, also defined in \cite{Beers1990}, and $C$ is the a tuning constant. We choose $C=6$ (the suggested value) which balances computational speed and location accuracy. 
% Although this work assumes that we know each galaxy's redshift to infinite precision, in practice, we find a simple weighted mean provides a reliable estimate of $z_c$ when there are uncertainties on the individual galaxy redshifts.

@@

For the clusters making up the \emph{test} sample the mass is unknown (it is what we are trying to predict) but the other features are known. To determine the mass probability distribution of a test cluster, $P(\log\ M_{200c})$, we combine the conditional probability distribution, $P(\log\ M_{200c}|\vec{x})$, created previously with the probability distribution of $\sigma$, the LOSVD, through Equation~\ref{eq: Pm}.
\begin{equation}\label{eq: Pm}
	P(\log\ M_{200c}) = \int P(\log\ M_{200c}|\vec{x})\ P(\log\ \sigma, z)\ d\log\ \sigma\ dz
\end{equation}
The expected mass is determined by calculating the first moment of the probability density. This becomes our ``predicted'' cluster mass, $M_{pred}$.
\begin{equation}\label{eq: expected mass}
	\log\ M_{pred}= \int \log\ M_{200c}\ P(\log\ M_{200c})\ d\log\ M_{200c}
\end{equation}
The confidence interval associated with this prediction can be estimated two ways. First, by calculating the second moment of the probability density through
\begin{equation}\label{eq: variance}
	\mathrm{Var} = \int (\log\ M_{200c} - \log\ M_{pred})^2\ P(\log\ M_{200c})\ d\log\ M_{200c}
\end{equation}
or by drawing many samples from $P(\log\ M_{200c})$ and calculating the values at the 16th and 84th percentile. In practice we find that both methods produce similar results for a large number of trials. Therefore, we quote predicted masses as the most probable mass given by Equation~\ref{eq: expected mass} and associated 68\% error estimated through the square root of Equation~\ref{eq: variance}.

@@

Because of the simplifying assumptions, and the superior quality of the data (no contamination, signal-to-noise issues, etc.), realistic expectations for HETDEX is to directly constrain $\sigma_8$ is not yet competitive with other methods (\eg, CMB, WL, X-ray). For example, \cite{DeHaan2016} constrain $\sigma_8$ to $\sim5$\% using a sample of 337 SZE detected clusters from the SPT-SZE survey.

We can use the rule-of-thumb calculations given in \citetalias{Weinberg2013} (Equations 154--155) to estimate the uncertainties required in both the absolute cluster mass scale and the intrinsic scatter to keep a cosmological analysis limited by cluster statistics. For the $\sim670$ clusters, with masses above $10^{14}$ \Msol, detected with Survey observations to constrain $\sigma_8$ and to be dominated by cluster statistics alone ($\Delta \ln N \sim N^{-1/2}$), the error on the  absolute cluster calibration would need to be better than 1.3\%. For a fully Targeted survey, about 1600 clusters with masses above $10^{14}$ \Msol, this cluster mass calibration uncertainty reduces to $<1\%$. Similarly, the uncertainty associate with the scatter must be approximately 2.2\% and 1.4\% for Survey and Targeted observations respectively, after assuming an intrinsic scatter of 20\%. Comparing these estimates to the uncertainties associated with the Targeted and Survey observations, we are dominated by the uncertainty associated with the absolute cluster mass scale calibration. Nevertheless, while analysis of HETDEX clusters may not provide the most accurate constraint on $\sigma_8$, it will provide important systematics checks on other studies and will ultimately improve the measurements of $\sigma_8$.

@@

We find that we are able to accurately recover an average intrinsic scatter of $0.2 <\langle \sigma_{M|\lambda} \rangle <0.3$ dex, finding $\langle \sigma_{M|\lambda} \rangle = 0.257\pm0.007$ at $\sigma_{true} = 0.25$ with Survey observations. As the intrinsic scatter increases or decreases, we fail to recover the scatter as accurately.

For the richness range $10 \leq \lambda < 130$ the intrinsic scatter in between the $\mathrm{M_{pred,corr}}$ predicted cluster masses and the true cluster mass (the basis of our richnesses) is $\sim0.15$ dex for Targeted observations and $\sim0.20$ dex for Survey observations. As we reduce the overall scatter in our cluster mass recovery, this floor will lower. We underestimate the scatter in high intrinsic scatter relations because any residual bias remaining in the predicted clusters masses after correction reduces the observed scatter. Because the bias subtraction when creating $\mathrm{M_{pred,corr}}$ subtracts the mean bias, we are left with a small amount of residual bias, lowering the measured scatter for high scatter relationships.

\section{Summary}\label{sec:summary}
Here, we present detailed simulations of the upcoming HETDEX survey's applicability to the detection and total mass measurement of galaxy clusters based on emission line and continuum redshifts for galaxies at z < 0.5. Using mock galaxy catalogs and simple HETDEX-like observational strategies and limits, we observe our simulated sky, estimate the number of clusters observed, and derive basic cluster parameters, redshift, line-of-sight velocity dispersion (LOSVD). Using a traditional power law-based, velocity dispersion, scaling relation along with more advanced probability and machine learning (ML) techniques, we estimate the total mass for each cluster. We discuss each cluster mass estimate's precision, and discuss the ability of HETDEX to constrain the cosmological parameter $\sigma_8$ based on those predicted cluster masses. In addition, we comment on how HETDEX may improve current and future photometric large-area sky surveys' cluster mass estimates derived from optical richness.
